### 2023-02-21 doc/manuscript/translocation.qmd
Edits throughout introduction and results

Emerging infectious diseases pose a severe threat to human, livestock, and wildlife populations. In wildlife populations, disease has been responsible for dramatic declines, extirpations, and even extinctions in mammals, birds, echinoderms, amphibians, and other taxa [@hewson2014; @samuel2015; @scheele2019; @cunningham2021]. Although the dramatic and immediate nature of disease-induced declines in wildlife has led to intensive focus on documenting, understanding, and mitigating disease impacts [@langwig2015], populations previously decimated by disease are also beginning to recover [@newell2013; @voyles2018]. Given the nascent nature of wildlife recoveries in the presence of disease, we know far less about the mechanisms driving population recovery [@brannelly2021; @russell2020] and how these mechanisms can be leveraged to re-establish host populations across large spatial scales [@mendelson2019]. In particular, for species of conservation concern as a result of disease-induced declines, demonstrating the long-term viability and feasibility of potential recovery actions in the presence of disease and elucidating the mechanisms underlying long-term recovery success remain largely absent from conservation biology [@venesky2012; @mendelson2019].

Host evolution following disease-induced declines is one mechanism by which hosts can recover [@carlson2014; @searle2020]. Evolutionary rescue describes natural host evolution whereby rapid evolutionary changes originating from variation within a host population reverse ongoing population declines [@carlson2014]. In host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021]. Evolutionary changes following disease-induced declines have been repeatedly observed in wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], but linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, remains rare. Important for conservation action, selection that promotes recoveries of focal populations can also provide sources of resistant individuals to reestablish populations extirpated following disease emergence [@joseph2018; @mendelson2019]. These so-called "recovery engines" [@mendelson2019] -- where rescued (i.e., resistant and recovered) host populations are used to successfully repatriate extirpated populations on the landscape -- are an intriguing management idea, but have not been successfully implemented in wildlife conservation efforts. <!--# Roland says: Modestly speaking, Joseph and Knapp (2018) does provide an example, and one that predates Mendelson (2019) paper. -->

<!--# Example text with more specific examples: black-lipped abalone populations suffering from herpes virus [@holland2022], little brown bat populations suffering from white-nose syndrome [@gignoux-wolfsohn2021], Tasmanian devils suffering from facial tumor disease [@epstein2016], and amphibian populations suffering from chytridiomycosis [@savage2016]-->

<!--# P3: In this paragraph, I took a broader approach than what was originally outlined (which focused on fitness benefits extending beyond natal environment). I'm happy to revisit this approach as necessary. Also, can we use "reduced susceptibilty" strategically (especially in Results) to avoid potential pitfalls associated with resistance, tolerance, and our lack of understanding of what actual mechanism underlies lower Bd loads and higher frog survival?  -->

@@

<!--#  Suggested start to this paragraph to link with the previous paragraph: "There are four conditions that will often..." -->

There are four conditions that will often determine the feasibility of this much-needed recovery approach. First, populations that survived following disease arrival have evolved increased resistance. Second, increased resistance contributes to higher individual fitness in the presence of disease and allows positive population growth. Third, the fitness benefits gained through selection for increased resistance manifest beyond the natal environment, allowing for the establishment of populations reintroduced into novel environments. And fourth, reintroduced populations show long-term growth and persistence in a landscape of disease. To date, few of these conditions are documented for any single species impacted by disease, and as a result the reintroduction of resistant hosts as a means to recover species of conservation concern remains largely an untested idea [@scheele2021]. Furthermore, a lack of relevant information about disease or a lack of suitable options to overcome the threat of disease has resulted in the failure of numerous reintroduction efforts [@scheele2021; @fellers2007]. With the increase in emerging infectious diseases in recent decades [@daszak2000], these failures highlight the critical need to develop effective tools to recover disease-impacted biodiversity worldwide. <!--# Need to better tie these two sentences to the conditions listed above. -->

<!--# P4: Description of MYL frog species complex, impacts of Bd, and evidence of recent recovery. -->

The mountain yellow-legged (MYL) frog, comprised of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by the recently-emerged amphibian chytrid fungus (*Batrachochytrium dendrobatidis*; [@scheele2019]). Once the most common amphibian in the high elevation portion of the Sierra Nevada mountains (California, USA; [@grinnell1924]), this frog has disappeared from more than 90% of its historical range during the past century [@vredenburg2007]. Due to the severity of its decline and the increasing probability of extinction, both species are now listed as "endangered" under the U.S. Endangered Species Act. Although this decline was initiated by the introduction of non-native trout into the once-fishless portion of the Sierra Nevada [@bradford1989; @knapp2000], the arrival of *B. dendrobatidis* (Bd) in the 1960s and its subsequent spread [@vredenburg2019] caused additional widespread population extirpations [@vredenburg2010; @rachowicz2006]. These Bd-caused declines are fundamentally different from the fish-caused declines because removal of fish populations is feasible and results in the rapid recovery of frogs. In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts [@walker2007], and therefore represents a permanent alteration of invaded ecosystems that amphibians will need to overcome to reestablish persistent populations.

<!--# P5: Here, we combine 24 translocations of resistant MYL populations, X longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate that resistant populations of an endangered amphibian species can recover and persist across a landscape of disease. We then use genetic evidence to demonstrate that observed disease resistance is likely a result of selection. As such, we conclude that natural selection combined with translocations promotes the recovery of a species on the brink of disease-induced extinction.  -->

Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some frog populations have persisted following epizootics [@briggs2010] and are now recovering [@knapp2016]. Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant genotypes [@knapp2016]. Furthermore, exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated indicate that population establishment is possible despite ongoing Bd infection [@joseph2018]. Here, we report the results of a greatly expanded MYL frog reintroduction effort and an associated analysis to evaluate genomic signatures of selection following MYL frog exposure to Bd. Collectively, the results show that MYL frogs generally meet the four conditions for reintroduction feasibility outlined above, and provide a novel example of how the reintroduction of resistant frogs can allow the large-scale recovery of this disease-impacted amphibian. These results may have broad implications for the many species worldwide that are threatened with extinction by emerging diseases.

@@

#### Translocations allow population reestablishment

@@

<!--# Likely need an introductory few sentences to put the translocations in the proper context. -->

@@

Reintroductions of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species. These reintroductions involved the collection of adult frogs from nearby large donor populations that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to recipient sites. During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of frogs to `r n_distinct(d1$site_id)` recipient sites (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap) <!--# How to restart numbering for figures in SI, add "S" prefix? -->. Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs collected from one of `r n_distinct(d1$donor)` donor populations (@fig-yosemap). Estimates of frog survival during the one year following translocation indicate that survival was highly variable between sites, but relatively constant within sites (for the subset of sites that received multiple translocations; @fig-translocation-survival). As such, for sites that received more than one translocation, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in all subsequent translocations to that site (median of posterior distribution = 4.14, uncertainty interval = 3.06\--5.28; see Supporting Information for analysis details). The relatively constant within-site frog survival across all translocations suggests an important effect of site characteristics on frog survival. <!--# How to extract these values from table produced by line 934 in translocation_survival_analysis.Rmd (or from associated code output)? -->

The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens). This result is consistent with our hypothesis that frogs in recovering populations have evolved increased resistance and/or tolerance to Bd infection. Important predictors are shown in @fig-cond-effects, and included winter severity in the year following translocation (snow_t1), site elevation (elevation), and donor population (donor_72996). Males had somewhat higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects). In addition to Bd load, effects of frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens).

<!--# Suggested text to wrap this section up and build into the next two sections: Overall, our analyses show that putatively resistant frogs can maintain the benefits of resistance (i.e., minimal Bd-induced mortality despite ongoing infection) in non-natal habitats. Moreover, in three locations where longer mark-recapture time series allow us to assess the survival of adults recruited to the population due to the breeding of translocated individuals, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults. These results show that putative resistance in non-natal habitats is maintained across generations of frogs, suggesting that acquired, heritable resistance in donor populations could promote the long-term viability of translocated populations.  We test these claims in the following sections. -->

@@

<!--# Add concluding 1-2 sentences to emphasize take-home message. -->

<!--# Present results indicating that Bd loads on frogs changed little following translocation? -->

@@

<!--# Likely need an introductory few sentences to put the viability analyses in the proper context.  Mark W: I have made an attempt at this. -->

@@

>>>>

Emerging infectious diseases are increasing [@daszak2000] and pose a severe threat to human, livestock, and wildlife populations. In wildlife populations, disease is responsible for dramatic declines, extirpations, and even extinctions in mammals, birds, amphibians, echinoderms, and other taxa [@hewson2014; @samuel2015; @scheele2019; @cunningham2021]. Although the dramatic and immediate nature of disease-induced declines in wildlife has led to intensive focus on documenting and understanding disease impacts [@langwig2015], this information is generally insufficient to allow the development of effective methods to recover disease-impacted species [@scheele2021; @fellers2007]. Importantly, some populations previously decimated by disease have subsequently recovered to varying degrees [@newell2013; @voyles2018; @knapp2016] and might provide critical insights for species conservation efforts. Unfortunately, given the nascent nature of wildlife recoveries in the presence of disease, we know far less about the mechanisms driving population recovery [@brannelly2021; @russell2020] and how these mechanisms can be leveraged to re-establish host populations across large spatial scales [@mendelson2019]. At a time when novel infectious diseases are emerging at an unprecedented rate, this knowledge gap has far-reaching consequences for global biodiversity.

Host evolution following disease-induced declines is one mechanism by which hosts can recover [@carlson2014; @searle2020]. Evolutionary rescue describes natural host evolution whereby rapid evolutionary changes originating from variation within a host population reverse ongoing population declines [@carlson2014]. In host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021]. Evolutionary changes following disease-induced declines are known from a diversity wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], but linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, remains rare. Important for conservation action, selection that promotes recoveries of focal populations can also provide sources of individuals with increased resistance and/or tolerance (hereafter, "resistance") to reestablish populations extirpated following disease emergence [e.g., @mendelson2019]. To date, this potentially important species recovery strategy is rarely utilized in wildlife conservation efforts [see @joseph2018 for a notable example], and studies that evaluate its potential utility are critically needed. <!--# Roland says: Modestly speaking, when combined with the selection results in the current paper, Joseph and Knapp (2018) does provide an example. -->

There are four conditions that will often determine the feasibility of using individuals from populations that are recovering following disease-induced declines to reestablish extirpated populations. First, populations that survived following disease arrival have evolved increased resistance. Second, increased resistance contributes to higher individual fitness in the presence of disease and allows positive population growth. Third, the fitness benefits gained through selection for increased resistance manifest beyond the natal environment, allowing for the establishment of populations reintroduced into novel environments. And fourth, reintroduced populations show long-term growth and persistence in a landscape of disease. To date, few of these conditions are documented for any single species impacted by disease, and as a result the reintroduction of resistant hosts as a means to recover species of conservation concern remains largely an untested idea [@scheele2021].

The mountain yellow-legged (MYL) frog, comprised of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by the recently-emerged amphibian chytrid fungus (*Batrachochytrium dendrobatidis*; [@scheele2019]). Once the most common amphibian in the high elevation portion of the Sierra Nevada mountains (California, USA; [@grinnell1924]), this frog has disappeared from more than 90% of its historical range during the past century [@vredenburg2007]. Due to the severity of its decline and the increasing probability of extinction, both species are now listed as "endangered" under the U.S. Endangered Species Act. Although this decline was initiated by the introduction of non-native trout into the once-fishless portion of the Sierra Nevada [@bradford1989; @knapp2000], the arrival of *B. dendrobatidis* (Bd) in the 1960s and its subsequent spread [@vredenburg2019] caused additional widespread population extirpations [@vredenburg2010; @rachowicz2006]. These Bd-caused declines are fundamentally different from the fish-caused declines because removal of fish populations is feasible and results in the rapid recovery of frogs [@knapp2007; @vredenburg2004]. In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts [@walker2007], and therefore represents a permanent alteration of invaded ecosystems that amphibians will need to overcome to reestablish persistent populations.

Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some frog populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016]. Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant genotypes [@knapp2016]. Furthermore, exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated indicate that population establishment is possible despite ongoing Bd infection [@joseph2018]. Here, we report the results of a greatly expanded MYL frog reintroduction effort and an associated analysis to evaluate genomic signatures of selection following MYL frog exposure to Bd. Collectively, the results show that MYL frogs generally meet the four conditions for reintroduction feasibility outlined above, and provide a novel example of how the reintroduction of resistant frogs can allow the large-scale recovery of this disease-impacted amphibian. These results may have broad implications for the many species worldwide that are threatened with extinction by emerging diseases.

@@

@@

<!--# Let's revisit whether the selection results should go here or at the end of the results. Putting those results here would reduce the need to describe populations as "putatively resistant". -->

#### Reintroductions allow population reestablishment

@@

@@

Reintroductions of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species. These reintroductions involved the collection of adult frogs from large, Bd-positive donor populations <!--# Add figure to SI showing Bd loads before and after translocations -->that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to nearby recipient sites. We refer to these reintroductions as "translocations". During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of frogs to `r n_distinct(d1$site_id)` recipient sites (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap) <!--# How to restart numbering for figures in SI, add "S" prefix? -->. Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs collected from one of `r n_distinct(d1$donor)` donor populations (@fig-yosemap). Estimates of frog survival during the one year following translocation indicate that survival was highly variable between sites, but relatively constant within sites (for the subset of sites that received multiple translocations; @fig-translocation-survival). As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in all subsequent translocations to that site (median of posterior distribution = 4.14, uncertainty interval = 3.06\--5.28 <!--# How to extract these values from table produced by line 934 in translocation_survival_analysis.Rmd (or from associated code output)? --> ; see Supporting Information for analysis details), and suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival. For frogs translocated to each site, Bd load changed little between the periods immediately prior to translocation versus during the one year after translocation (see SI for analysis details).

To identify drivers of post-translocation frog survival, we developed a hierarchical model in a Bayesian framework (see SI for details). Our specific interest was whether Bd load had an important effect on adult survival, as would be expected if the effect of frog resistance was insufficient to overcome the negative effect of Bd. The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens). Important predictors included winter severity in the year following translocation (snow_t1), site elevation (elevation), and donor population (donor_72996; @fig-cond-effects). Males had somewhat higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects). In addition to Bd load, effects of frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens). In summary, our analyses show that Bd load is not a predictor of the survival of translocated frogs, consistent with frogs in recovering populations having evolved increased resistance and/or tolerance to Bd infection. In addition, the small changes in Bd load following translocation and relatively high survival in the majority of translocated cohorts suggests that putatively resistant frogs from recovering populations can maintain the benefits of resistance (i.e., low-moderate Bde loads, minimal Bd-induced mortality despite ongoing infection) in non-natal habitats. Finally, in three locations where longer mark-recapture time series allowed us to assess the survival of adults recruited to the population that are the progeny of translocated individuals, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults <!--# Show results. -->. Collectively, these results show that putative resistance in non-natal habitats is maintained across generations of frogs.

@@

@@

@@

![For frogs translocated to each recipient site, change in Bd load between the periods immediately prior to translocation versus during the 1-year period after translocation.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="450"}
